---
title: "Beaked whale click detection"
format: html
editor: visual
---

## Looking for clicks

```{r}
#| message: false
#| warning: false

library(tuneR)
library(tidyverse)
library(signal)
library(oce)
```

I downloaded a 10 minute recording taken by the MARS hydrophone on June 13, 2018 around 2:30 PM. I chose this date because we know beaked whales were in Monterey Bay at this point.

```{r}

#Read in the data  
data <- readWave("data/MARS_20180613_142912.wav")
sound <- data@left
duration <- length(sound) / data@samp.rate
fs <- data@samp.rate

#Write a function to clip the sound
clip_by_sec <- function(x, f, start, end) {
  n_x <- as.numeric(length(x))
  f <- as.numeric(f)
  stopifnot(length(start) == 1,
            length(end) == 1,
            start >= 0,
            end >= 0,
            start <= length(x) * f,
            end <= length(x) * f)
  start_idx <- 1 + start * f
  end_idx <- 1 + end * f
  x[start_idx:end_idx]
}
```

I have a .mat file from Brent with tons of clicks. I wanted to cross reference the clicks with the MARS hydrophone data on a day that we know beaked whales were around. Next we clipped the first half a second of the .wav file and overlay the click locations on top.

```{r}

#Get the first half a second
sound2 <- clip_by_sec(sound, fs, 0, .5)
fs2 <- fs 

#Create spectrogram
spec <- specgram(x = sound2, 
                 n = 1024, 
                 Fs = fs2, 
                 window = 256, 
                 overlap = 128)
#discard phase information
p <- abs(spec$S)
#normalize
p <- p / max(p)
#config time axis
t <- spec$t

# zoom in on 20 - 120 kHz
flim <- c(20e3, 120e3)
plim <- range(p[between(spec$f, flim[1], flim[2]), ])
imagep(x = t, 
       y = spec$f, 
       z = t(p), 
       col = oce.colorsViridis, 
       ylab = "Frequency", 
       xlab = "Time", 
       #xlim = c(80, 90),
       ylim = flim,
       zlim = plim,
       drawPalette = TRUE, 
       decimate = FALSE)
#Manually show where the clicks are (from Brent's .mat file)
clicks <- c(0.0341, 0.0776, 0.0901, 0.1446, 0.1889, 0.1942, 0.1987, 0.2514)
abline(v = clicks, col = "red")
```

The clicks appear to be correct. Now, we try to calculate some of the features of the clicks that we could use to identify the species, starting with peak and center frequency.

```{r}
#read in the .mat file 
clicks_mat <- R.matlab::readMat("data/training/yearly/2018/clicks/MARS_20180613_142912.wav_clicks.mat")
```

```{r}

# Signal frequency power 

# First: high pass filter above 10^3.5 Hz
nyquist <- fs / 2
W <- c(10^4, 10^5) / nyquist
bf <- butter(4, W, type = "pass")
sound_high <- filtfilt(bf, sound)

#pick a click to try to describe the features
myclick_t <- mean(c(clicks_mat$start[100], clicks_mat$stop[100]))

window_size_s <- 0.1
window_size_i <- window_size_s * fs
myclick_win <- as.integer(myclick_t * fs + c(-1, 1) * window_size_i / 2)
myclick_signal <- sound_high[myclick_win[1]:myclick_win[2]]
myclick_fft <- fft(myclick_signal)
n <- length(myclick_signal)
f <- (0:(n - 1)) * (fs / n)
power = abs(myclick_fft)^2 / n
in_range <- f >= 10^4 & f <= 10^5
plot(f[in_range], power[in_range], type = "l")

#But these seem wrong
peak_freq <- f[which.max(power)]
center_freq <- weighted.mean(f, power)
peak_freq 
center_freq
```

The the peak and center frequency we calculated doesn't seem to be correct. Back to this [tutorial](https://www.mathworks.com/help/matlab/math/basic-spectral-analysis.html) to try again?
