---
title: "Beaked whale click detection"
format: html
editor: visual
---

### Classification

------------------------------------------------------------------------

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(dbscan)
```

Attempting to translate Sam's Julia code into R.

First up: read, sample, and standardize the features

```{r}
#read in a feather file. every feather file holds a day.
clicks <- dir(here::here("data"), pattern = "features_.*feather", full.names = TRUE) %>% 
  map(arrow::read_feather) %>% 
  list_rbind() %>% 
  #selecting the four features that Sam used to cluster: duration, peak frequency, peak bandwidth, and median bandwidth, and transforming duration to log(duration) 
  mutate(logduration = log(duration)) %>% 
  select(logduration, peak_freq, peak_bw, freq50)
#  select(logduration, peak_freq, freq50)

# sample 10k clicks
set.seed(1407)
click_sample <- slice_sample(clicks, n = 1e4)

#normalize everything by z score (sam did along 0-1)
#z score better when data is sensitive to outliers
#0-1 better when data distribution is unknown or heavily skewed...
z_score <- function(x) {
  stopifnot(sum(!is.na(x)) > 3)
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}
click_norm <- mutate(click_sample, across(everything(), z_score))


```

HDBSCAN is a *neighborhood based* clustering algorithm. It takes one parameter, `minPts,` which sets a minimum for cluster size. Let's set `minPts` to 100.

```{r}
click_minPts <- 5
hdb_clicks <- hdbscan(click_norm, minPts = click_minPts)
plot(hdb_clicks)
```

```{r}
foo <- kNN(click_norm, k = 10)
plot(foo, click_norm)
#cl <- optics(click_norm, minPts = 10)
```

What do the normalized click features look like in space? Use a pairs plot to visualize.

```{r}
library(GGally)
ggpairs(click_norm, mapping = aes(color = "color", alpha = "alpha")) +
  scale_color_manual(values = "cornflowerblue") +
  scale_alpha_manual(values = 0.25) +
  theme_bw()
```

Reduce from 4 dimensions to 2 using UMAP

```{r}
library(umap)
click_umap <- umap(click_norm, n_components = 2)
click_umap_df <- as.data.frame(click_umap$layout)
colnames(click_umap_df) <- c("umap1", "umap2")

archetypes_umap <- click_umap_df[(nrow(archetypes) - 3):nrow(archetypes), ]

click_umap_hdb <- hdbscan(click_umap_df, minPts = 50)
plot(click_umap_hdb)

click_umap_df <- click_umap_df %>% 
  mutate(cl = click_umap_hdb$cluster,
         prob = click_umap_hdb$membership_prob)

ggplot(click_umap_df, aes(umap1, umap2)) +
  geom_point(data = select(click_umap_df, -cl), color = "grey70", alpha = 0.1, size = 0.5) +
  geom_point(aes(color = factor(cl), alpha = prob), size = 0.8) +
  facet_wrap(~factor(cl)) +
  theme_classic() +
  theme(legend.position = "none")
ggsave(here::here("figs/umap_hdbscan.pdf"), height = 6.5, width = 9)

```

```{r}

click_umap <- umap::umap(click_norm, n_components = 2)

click_umap_df <- as.data.frame(click_umap$layout)
colnames(click_umap_df) <- c("umap1", "umap2")
minPts <- 25
kNNdistplot(click_umap_df, k = minPts - 1)
# Find the "knee" by eyeballing the (k-1)NN plot
click_eps <- 0.9
abline(h = click_eps, col = "firebrick", lty = 2)

```

```{r}

click_umap_dbscan <- dbscan(click_umap_df, eps = click_eps, minPts = minPts)
click_umap_dbscan

```

```{r}

click_umap_df$cl <- factor(click_umap_dbscan$cluster)

ggplot(click_umap_df, aes(umap1, umap2)) +
  geom_point(data = select(click_umap_df, -cl), color = "grey70", alpha = 0.1, size = 0.5) +
  geom_point(aes(color = cl), size = 0.8) +
  facet_wrap(~cl) +
  theme_classic() +
  theme(legend.position = "none")
ggsave(here::here("figs/umap_dbscan.pdf"), height = 6.5, width = 9)

```

match filter for echosounder clicks - 38 kHz

or focus on a time when there is no echosounder and there are beaked whales. check the visual browser.

```{r}
archetypes <- read_csv(here::here("data/click_archetypes.csv")) %>% 
  mutate(duration = duration / 1000, # convert ms to s
         peak_freq = peak_freq * 1000,
         freq50 = freq50 * 1000,
         logduration = log(duration)) %>% 
  select(logduration, peak_freq, freq50)

click_sample <- click_sample %>% rbind(archetypes)
```
