---
title: "Beaked whale click detection"
format: html
editor: visual
---

### Classification

------------------------------------------------------------------------

```{r}
#| message: false
#| warning: false

library(arrow)
library(tidyverse)
library(dbscan)
# see neonscience.org/resources/learning-hub/tutorials/hdf5-intro-r
library(rhdf5)
```

Attempting to translate Sam's Julia code into R.

First up: read, sample, and standardize the features

```{r}
#read in a feather file
clicks <- arrow::read_feather(here::here("data/features_20190529.feather"))
#every feather file holds a day

#transforming duration to log(duration) and selecting the four features that Sam used to cluster
clicks <- clicks %>% 
  mutate(logduration = log(duration)) %>% 
  select(logduration, peak_freq, peak_bw, freq50)

# sample 10k clicks
set.seed(1407)
click_sample <- slice_sample(clicks, n = 1000)
# click_idx <- sample(1:nrow(clicks), 22200)
# click_sub <- clicks[click_idx, ]
#click_sub <- clicks

#normalize everything along 0-1
# click_norm_sub <- (click_sub - min(click_sub)) / (max(click_sub) - min(click_sub))
z_score <- function(x) {
  stopifnot(sum(!is.na(x)) > 3)
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}
click_norm <- mutate(click_sample, across(everything(), z_score))


```

DBSCAN is a *neighborhood based* clustering algorithm. It takes two parameter, `eps`, which weights the neighborhoods (?) and `minPts` which sets a minimum for cluster size. Let's set `minPts` to 100. Then we use the k-Nearest Neighbors distance plot to determine a good choice for `eps`.

```{r}
click_minPts <- 10
click_kNNdist <- kNNdist(click_norm, k = click_minPts - 1)
# Used identify() to figure out the knee indices
knees_idx <- c(107, 922)
kNNdistplot(click_norm, k = click_minPts - 1)
abline(h = c(sort(click_kNNdist)[knees_idx]), lty = 2, col = "firebrick")
```

From the plot above we have two different candidates for `eps`: 0.1743 or 1.766.

```{r}
click_cluster1 <- dbscan(click_norm, eps = 0.1743, minPts = click_minPts)
click_cluster1
```

```{r}
click_norm %>% 
  mutate(cluster = factor(click_cluster1$cluster)) %>% 
  GGally::ggpairs(aes(color = cluster), columns = 1:4)
```

The first candidate for `eps` gives us nonsense results because there's a hyper-dense cluster that is causing a false knee. Let's try the larger `eps` candidate.

```{r}
click_cluster2 <- dbscan(click_norm, eps = 1.766, minPts = click_minPts)
click_cluster2
```
